---
title: "Practical Machine learning write up"
author: "Roshni Sharma"
date: "July 12, 2018"
output:
  pdf_document: default
  html_document: default
---

This is a write up for Prediction Assignment from Coursera for Practical Machine learning course.Background, data and Goal section is directly taken from the question.

# Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# Data

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


# Goal
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

## Model creation and selection

### Understanding data from stated site:
Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: 
exactly according to the specification (Class A), 
throwing the elbows to the front (Class B), 
lifting the dumbbell only halfway (Class C), 
lowering the dumbbell only halfway (Class D) and 
throwing the hips to the front (Class E).
Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. 

### Model selection approach:
I am going to start with decision tree and then random forest. If both these do not give good accuracy then i will try model based prediction or by combining predictors.

### Packages, Libraries, Reproducibility
Installing packages, loading libraries, and setting the seed for reproduceability

```{r}
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(rattle)
set.seed(12345)
```
###Load data
```{r}
trainingdata <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"), na.strings=c("NA","#DIV/0!",""))
testdata <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"), na.strings=c("NA","#DIV/0!",""))
```
###Create training, test and validation sets
```{r}
inTrain <- createDataPartition(y=trainingdata$classe, p=0.6, list=FALSE)
myTraining <- trainingdata[inTrain, ] 
myTesting <- trainingdata[-inTrain, ]
dim(myTraining)
#Training set has 13737 rows and 160 features.
```
###Prepocessing
#####Subsetting traning data for cross validation
```{r}
train_set <- createDataPartition(y= myTraining$classe, p = 0.7, list = FALSE)
Training <- myTraining[train_set, ]
Validation <- myTraining[-train_set, ]
dim(Training); dim(Validation)
```
###Feature selecting
####Removing features with not much variability or impact on class, or missing values, or descriptive values which cannot contribute to the model
```{r}
nzvnames<- nearZeroVar(Training)
Training <- Training[, -nzvnames] 
countna <- sapply(Training, function(x) {
  sum(!(is.na(x) | x == ""))
  })
 nullcol <- names(countna[countna < 0.8 * length(Training$classe)]) 
 nocontricol <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", 
    "cvtd_timestamp", "new_window", "num_window")
 excludecolumns <- c(nullcol,nocontricol)
 Training <- Training[, !names(Training) %in% excludecolumns]
 dim(Training); head(Training)
 #doing same transformation to test and Validation data
 nzvnames<- nearZeroVar(Training)
 colnametraining <- colnames(Training)
Validation <- Validation[, colnametraining] 
myTesting <- myTesting[, colnametraining] 
```
####Ploting predictors starting with total
```{r}
featurePlot(x=Training[,c("total_accel_belt" ,"total_accel_arm", "total_accel_dumbbell","total_accel_forearm","total_accel_belt")],y = Training$classe,plot="pairs")
```
##Creating Machine learning predictions - Decision Tree model1
```{r}
model1 <- rpart(classe ~ ., data=Training, method="class")
fancyRpartPlot(model1)
```
###Predict
```{r}
predmodel1in <-predict(model1, Training, type = "class")
predmodel1 <- predict(model1, Validation, type = "class")

```
###Accuracy of MOdel1

```{r}
# insample accuracy
confusionMatrix(predmodel1in, Training$classe)
# out sample accuracy
confusionMatrix(predmodel1, Validation$classe)
```

##Creating Machine learning predictions - RandomForest model2
```{r}
model2 <- randomForest(classe ~. , data=Training)
```
###Predict
```{r}
predmodel2in <-predict(model2, Training, type = "class")
predmodel2 <- predict(model2, Validation, type = "class")

```
###Accuracy of MOdel2

```{r}
# insample accuracy
confusionMatrix(predmodel2in, Training$classe)
# out sample accuracy
confusionMatrix(predmodel2, Validation$classe)

# Clearly Random FOrest has much better accuracy (of 100% for insample and 98.9% for outsample)
```
###FIles for submission
```{r}
predmodel2test <- predict(model2, myTesting, type = "class")
```
Function to generate files with predictions to submit for assignment
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predmodel2test)
